---
title: "load-data"
author: "Grayson White"
date: "10/30/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Loading the GSS dataset

Note: Do not run this chunk. Only needed to initially pull the data from the GSS database. We have now saved this as `gss_orig.rda` and it is loaded by `load("gss_orig.rda")` in the chunk after this.
```{r eval = FALSE}
temp <- tempfile()
download.file("https://gss.norc.org/documents/stata/GSS_stata.zip",temp)

gss_orig <- haven::read_dta(unz(temp, filename = "GSS7218_R3.DTA")) %>%
  haven::as_factor()
unlink(temp)

save(gss_orig, file = "gss_orig.rda")
```

## Filtering

Note: Most of this filtering was done for the `infer` package `gss` dataset and can be attributed to authors of that package. We have included more rows and columns than that package, however, much initial tidying and subsetting can be attributed to them. (Make sure to cite this in the final document)
```{r}
load("gss_orig.rda")

gss_subset <- gss_orig %>%
  filter(!stringr::str_detect(sample, "blk oversamp")) %>% # this is for weighting
  select(
    year,
    age,
    sex,
    college = degree,
    partyid,
    hompop,
    hours = hrs1,
    income,
    class,
    finrela,
    wrkgovt,
    marital,
    educ,
    race,
    incom16,
    weight = wtssall
  ) %>%
  mutate_if(is.factor, ~ fct_collapse(., NULL = c("IAP", "NA", "iap", "na"))) %>%
  mutate(
    age = age %>%
      fct_recode("89" = "89 or older",
                 NULL = "DK") %>%
      as.character() %>%
      as.numeric(),
    hompop = hompop %>%
      fct_collapse(NULL = c("DK")) %>%
      as.character() %>%
      as.numeric(),
    hours = hours %>%
      fct_recode("89" = "89+ hrs",
                 NULL = "DK") %>%
      as.character() %>%
      as.numeric(),
    weight = weight %>%
      as.character() %>%
      as.numeric(),
    partyid = fct_collapse(
      partyid,
      dem = c("strong democrat", "not str democrat"),
      rep = c("strong republican", "not str republican"),
      ind = c("ind,near dem", "independent", "ind,near rep"),
      other = "other party"
    ),
    income = factor(income, ordered = TRUE),
    college = fct_collapse(
      college,
      degree = c("junior college", "bachelor", "graduate"),
      "no degree"  = c("lt high school", "high school"),
      NULL = "dk"
    )
  ) %>%
  filter(year >= 2000) %>%
  filter(partyid %in% c("dem", "rep")) %>%
  drop_na()

# save(gss_subset, file = "gss_subset.rda") # only run if need to save over the subset data. (Adding a column, or something)
```


