---
title: "Survey"
date: "12/3/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, we can load the data and tidy it: 
```{r}
library(tidyverse)
load("gss/gss_orig.rda")
gss_survey <- gss_orig %>%
  filter(!stringr::str_detect(sample, "blk oversamp")) %>% # this is for weighting
  select(vpsu, vstrat, year, age, sex, college = degree, partyid, hompop, hours = hrs1, income, class, finrela, wrkgovt, marital, educ, race, incom16, weight = wtssall) %>%
  mutate_if(is.factor, ~ fct_collapse(., NULL = c("IAP", "NA", "iap", "na"))) %>%
  mutate(
    age = age %>%
      fct_recode("89" = "89 or older",
                 NULL = "DK") %>%
      as.character() %>%
      as.numeric(),
    hompop = hompop %>%
      fct_collapse(NULL = c("DK")) %>%
      as.character() %>%
      as.numeric(),
    hours = hours %>%
      fct_recode("89" = "89+ hrs",
                 NULL = "DK") %>%
      as.character() %>%
      as.numeric(),
    weight = weight %>%
      as.character() %>%
      as.numeric(),
    partyid = fct_collapse(
      partyid,
      dem = c("strong democrat", "not str democrat"),
      rep = c("strong republican", "not str republican"),
      ind = c("ind,near dem", "independent", "ind,near rep"),
      other = "other party"
    ),
    income = factor(income, ordered = TRUE),
    college = fct_collapse(
      college,
      degree = c("junior college", "bachelor", "graduate"),
      "no degree"  = c("lt high school", "high school"),
      NULL = "dk"
    )
  ) %>%
  filter(year >= 2000) %>%
  filter(partyid %in% c("dem", "rep")) %>%
  drop_na()
```

Now, we construct a complex sample survey design. 
```{r}
library(survey)
gss_design <- 
    svydesign( 
        ~ vpsu , 
        strata = ~ vstrat , 
        data = gss_survey , 
        weights = ~ weight , 
        nest = TRUE 
    )
```


Now, we fit the logistic-regression model with weights using the svyglm() function from the survey package. A slight wrinkle is that we must use the quasibinomial rather than the binomial family to avoid a warning about noninteger counts produced by the use of differential sampling weights. 

```{r}
 options(survey.lonely.psu="certainty")
glm_result <- 
    svyglm( 
        partyid ~ age + sex + college + hompop + hours +
          income + class + finrela + wrkgovt + marital +
        educ + race + incom16 + weight, design=gss_design, family=quasibinomial)

summary(glm_result)
```

Note: The global option,  options(survey.lonely.psu="fail"), makes it an error to have a stratum with a single, non-certainty PSU. Changing it to options(survey.lonely.psu="certainty"), single-PSU stratum makes no contribution to the variance (for multistage sampling it makes no contribution at that level of sampling). This is an alternative to specifying fpc, and is useful to run the regression without error.

```{r}
probs_survey<-predict(glm_result, gss_survey, type = "response")
preds_survey<-ifelse(probs >=.5, 1, 0)
conf_log_survey <- table(preds, gss_survey$partyid)
conf_log_survey
n <- length(gss_survey$partyid)
false_pos_survey <- conf_log[1,2]
false_neg_survey <- conf_log[2,1]
error_survey <- 1/n *(false_pos + false_neg)
error_survey
```
We see that the error rate is 0.1810345 for the logistic regression with weights. 


