---
title: "Technical Report"
author: "Shisham Adhikari, Maggie Slein, Grayson White"
bibliography: references.bib
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(out.width = "75%")
knitr::opts_chunk$set(fig.align = "center")
library(tidyverse)
theme_set(theme_bw())
```

# Data 

## The GSS dataset

The general social survey (GSS) is a massive survey conducted of people within the United States since 1972. The GSS aims to get a representative sample of people in the United States and to understand information about them and how they feel about social and political issues. We have chosen some key variables collected from this survey, along with participants from 2000 or more recent, in order for us to attempt to classify political affiliation of participants. Our subset of the GSS dataset contains 5,800 rows, 16 columns, and 0 NA's.  

## Filtering

Most of this filtering was done for the `infer` package `gss` dataset and can be attributed to authors of that package. We have included more rows and columns than that package, however, much initial tidying and subsetting can be attributed to them [@infer]. Below is the code adapted from the `infer` package to attain our dataset, `gss_subset`:
```{r gss_filter, message = FALSE, eval = FALSE}
load("gss/gss_orig.rda")
gss_subset <- gss_orig %>%
  filter(!stringr::str_detect(sample, "blk oversamp")) %>% # this is for weighting
  select(year, age, sex, college = degree, partyid, hompop, hours = hrs1, income,
         class, finrela, wrkgovt, marital, educ, race, incom16, weight = wtssall) %>%
  mutate_if(is.factor, ~ fct_collapse(., NULL = c("IAP", "NA", "iap", "na"))) %>%
  mutate(
    age = age %>%
      fct_recode("89" = "89 or older",
                 NULL = "DK") %>%
      as.character() %>%
      as.numeric(),
    hompop = hompop %>%
      fct_collapse(NULL = c("DK")) %>%
      as.character() %>%
      as.numeric(),
    hours = hours %>%
      fct_recode("89" = "89+ hrs",
                 NULL = "DK") %>%
      as.character() %>%
      as.numeric(),
    weight = weight %>%
      as.character() %>%
      as.numeric(),
    partyid = fct_collapse(
      partyid,
      dem = c("strong democrat", "not str democrat"),
      rep = c("strong republican", "not str republican"),
      ind = c("ind,near dem", "independent", "ind,near rep"),
      other = "other party"
    ),
    income = factor(income, ordered = TRUE),
    college = fct_collapse(
      college,
      degree = c("junior college", "bachelor", "graduate"),
      "no degree"  = c("lt high school", "high school"),
      NULL = "dk"
    )
  ) %>%
  filter(year >= 2000) %>%
  filter(partyid %in% c("dem", "rep")) %>%
  drop_na()
```
```{r actually_load_gss, echo = FALSE}
# run this chunk to quickly load the gss subset without having to re-run the filtering
load("gss/gss_subset.rda")
```

Given our goal to understand which factors influence party affiliation in the US, we selected `year` (year of the election), `age` (age of time of survey), `college` (degree or no degree), `partyid` (democrat or republican), `hompop` (number of people in the respondent's household), `hours` (number of hours worked in the last week), `income` (total family income, categorical), `class` (socioeconomic class as described by respondent), `finrela` (respondent's opinion on family's income level), `wrkgovt` (whether or not the respondent works for the government), `marital` (respondent's martial status), `educ` (highest year of school completed), `race` (race of respondent), `income16` (respondent's family income at the age of 16), and `weight` (survey weight). 

We made some choices while filtering the dataset which will effect the final results of our models. First of all, we have filtered all observations which do not state that their political affiliation was either democrat or republican. We are most interested in answering the question of whether or not we can classify between these parties rather than considering much smaller third parties. Also, we have filtered all observations with any NA's. We chose to do this for ease of analysis and because many of the models we use will not consider a row that includes NA's in any of the columns being used for the model.  

# Exploratory Data Analysis

Before we dig too deeply in to the dataset, it is important to understand its structure: 
```{r}
# Number of rows
nrow(gss_subset) 

# Number of columns
ncol(gss_subset) 

# Response variable summary
summary(gss_subset$partyid)

# Data Structure
str(gss_subset)

# Glimpse of dataset
gss_subset %>%
  select(-weight) %>%
  rename(home = hompop, party = partyid) %>%
  head() %>%
  knitr::kable()
```

Below are plots that show the distribution of political party affiliation between democrat and republican as well as the distrutbution of all the predictors included in this dataset.

Most of our predictors appear to normally distributed, except for income, hompop, and weight. None of the predictors appear to have a strong relationship with political party affiliation, which is not surprising given that there are roughly the same amount of democrats and republicans in each state. 

First, we can look at the distribution of democrats and republications in our dataset in counts:
```{r partyid-dist, echo = FALSE}
ggplot(gss_subset, aes(x = partyid)) +
  geom_bar(fill = c("steelblue", "tomato"),
           color = "black") +
  labs(title = "Count of Democrat and Republican Party Affiliation in our Dataset")
```

There appears to me more democrats than republicans represented in this dataset, which could be because democrats are more likely to participate in this survey, or it could be that the way we selected our data systemically oversampled democrats. Notably from this, it is the case that our the weights associated with our sample of the GSS dataset would not be the same as the weights that the GSS uses for the dataset, so the `weight` variable should be ignored entirely. 

Now, we can examine some of our predictor variables with our response, `partyid`, to see the relationships there are between variables. First, we see in this side-by-side boxplot with the means plotted on top that republicans tend to be older on average:
```{r age-plot, echo = FALSE}
ggplot(gss_subset, aes(x = age,
                       y = partyid,
                       color = partyid)) +
  geom_boxplot(alpha = 0.3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  scale_color_manual(values = c("steelblue", "tomato")) +
  stat_summary(fun = mean, size = 0.25) +
  labs(title = "Political Affiliation by Age",
       x = "Age",
       y = "Political Affiliation",
       color = "Political Affiliation")
```

Next, it is interesting to consider economic status across political affiliations. By comparing political affiliation to income, class, and hours worked in the last week we can see small relationships between political affiliation and economic status:
```{r economic-status, echo = FALSE, warning = FALSE, message = FALSE}
ggplot(gss_subset, aes(x = income,
                       y = partyid,
                       color = partyid)) +
  geom_jitter(alpha = 0.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  scale_color_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Political Affiliation by Income",
    x = "Income",
    y = "Political Affiliation",
    color = "Political Affiliation"
  )

ggplot(gss_subset, aes(x = class,
                       y = partyid,
                       color = partyid)) +
  geom_jitter(alpha = 0.3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  scale_color_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Political Affiliation by Class",
    x = "Class",
    y = "Political Affiliation",
    color = "Political Affiliation"
  )

ggplot(gss_subset, aes(x = hours,
                       y = partyid,
                       color = partyid)) +
  geom_boxplot(alpha = 0.3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  scale_color_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Political Affiliation by Hours Worked Last Week",
    subtitle = "Average plotted with point on plot",
    x = "Hours Worked Last Week",
    y = "Political Affiliation",
    color = "Political Affiliation"
  ) +
  stat_summary(fun = mean, size = 0.25)
```


```{r}

```

```{r}

```


```{r, echo = FALSE}
ggplot(gss_subset, aes(x = race,
                       y = partyid,
                       color = partyid)) +
  geom_jitter(alpha = 0.3)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_manual(values = c("steelblue", "tomato"))

ggplot(gss_subset, aes(x = marital,
                       y = partyid,
                       color = partyid)) +
  geom_jitter(alpha = 0.3)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_manual(values = c("steelblue", "tomato"))


reps <- gss_subset %>%
  filter(partyid == "rep")
dems <- gss_subset %>%
  filter(partyid == "dem")
ggplot() +
  geom_point(reps,
             mapping = aes(x = marital,
                           y = age,
                           color = partyid),
             position = position_nudge(x = -0.1),
             alpha = 0.5) +
  geom_point(dems,
             mapping = aes(x = marital,
                           y = age,
                           color = partyid),
             position = position_nudge(x = 0.1),
             alpha = 0.5) +
  scale_color_manual(values = c("steelblue", "tomato"))

ggplot(gss_subset, aes(x = as.numeric(educ),
                       y = partyid,
                       color = partyid)) +
  geom_boxplot(alpha = 0.3)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_manual(values = c("steelblue", "tomato"))
```

# References